Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model

### 2025-10-27 paper/paper.qmd

---
title: Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
filters:
  - acronyms
acronyms:
  keys:
  - shortname: CDI
    longname: _Clostridium difficile_ infection
    markdown: longname
  - shortname: VRE
    longname: vancomycin-resistant _Enterococcus_
    markdown: longname
  - shortname: ESBL
    longname: extended-spectrum beta-lactamase
  - shortname: CPE
    longname: carbapenemase-producing Enterobacterales
  - shortname: WHO
    longname: World Health Organization
  - shortname: IPC
    longname: infection prevention and control
  - shortname: ABM
    longname: agent-based model
  - shortname: MRSA
    longname: methicillin-resistant _Staphylococcus aureus_
    markdown: longname
  - shortname: SARS-CoV-2
    longname: severe acute respiratory syndrome coronavirus 2
  - shortname: ABC
    longname: approximate Bayesian computation
  - shortname: IQR
    longname: interquartile range
  - shortname: FIFO
    longname: first-in, first-out
  - shortname: SD
    longname: system dynamics
  insert_loa: false
---

@@

\acr[case=sentence]{IPC} measures are designed to prevent the spread of these pathogens within healthcare settings. Commonly used interventions include hand hygiene, environmental cleaning, and the isolation of infected/colonised patients [@worldhealthorganizationGlobalReportInfection2022].
A key role of \acr{IPC} teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
In practice, an outbreak is declared after rates of infection have been compared to historical rates, and this is followed by a series of interventions to control the outbreak.
These interventions may include increased frequency and thoroughness of environmental cleaning, cohorting of infected/colonised patients (particularly into single-occupancy _en suite_ rooms), and increased hand hygiene compliance.
The impact of these interventions is often difficult to quantify.
Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable.
Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such as rates of antimicrobial usage, and to structural differences such as hospital layout.

@@

The use of computer simulation techniques such as \acr{ABM} can be used to model the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact. Agent-based modelling has been used to describe the transmission of \acr{MRSA} [@jaramilloAgentBasedModel2015] and \acr{SARS-CoV-2} [@kerrCovasimAgentbasedModel2021]. An advantage of \acr{ABM} over other simulation approaches is that models can be used to project the impact of interventions on transmission, prior to their implementation [@cassidyMathematicalModellingHealth2019]. This can be particularly useful for both planning and active management of outbreaks, where the impact of individual interventions is difficult to quantify [@cassidyMathematicalModellingHealth2019].

Here, we describe an agent-based model that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure. Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods. The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}.

@@

-   $r_0$ is the baseline/*de novo* risk of infection/colonisation,

-   $m$ is the increased (or decreased) relative risk of infection/colonisation due to antimicrobial therapy,

@@

-   $dS/dt$ is the rate of change of susceptible/non-infected inpatients,
-   $S$ is the number of susceptible inpatients,
-   $I$ is the number of infected inpatients,
-   $A$ is the rate of new admissions,
-   $p$ is the community colonisation rate,
-   $\beta$ is the transmission coefficient,
-   $r$ is the rate of random (*de novo*) infection,
-   $D$ is the rate of patient discharges.

@@

```{r fig-sd-model, include=TRUE}
#| echo: false
#| fig.cap: "System dynamics model. $S$ and $I$ are the number of susceptible and infected patients, respectively; $A$ and $D$ are the rates of admitted and discharged patients, respectively; $p$ is the community colonisation rate; $\\beta$ is the transmission coefficient; and $r$ is the rate of random (*de novo*) infection."
knitr::include_graphics("./images/sd_model.svg")
```

@@

$$
D_{S,t} = \Bigl \lfloor D_{\text{total}, t} \cdot \frac{S_t}{S_t + I_t} \Bigr \rceil,
$$

@@

$$
D_{I,t} = \Bigl \lfloor D_{\text{total}, t} \cdot \frac{I_t}{S_t + I_t} \Bigr \rceil = D_{\text{total}, t} - D_{S,t},,
$$

@@

A similar approach was used to generate outbreaks in the stochastic system dynamics model, with the difference that the parameters were changed at: 1) onset of outbreak; and 2) end of outbreak. The parameters that were changed were: 1) $\beta$ (transmission coefficient); and 2) $p$ (community colonisation rate). The parameter $r$ (rate of random infection) was kept constant throughout the simulation.

@@

Model input parameters were broadly grouped into: 1) toilet (e.g., number of times toilet used per day, level of contamination after use or cleaning); 2) admission (e.g., community colonisation rate, antimicrobial prescription rate); and 3) colonisation parameters (e.g., relative risk increase of antimicrobial therapy on colonisation/infection).

@@

| Name | Distribution | Parameters | Description | Reference |
|---------------|---------------|---------------|---------------|---------------|
| Baseline (pre-outbreak) antimicrobial prescription rate | Constant | 49.9% | Proportion of patients administered antimicrobials during admission | @versportenAntimicrobialConsumptionResistance2018 |
| Admission duration | Poisson | $\lambda = 4.6  \text{ days}$ |  | @levitkryankelixhauserastrangesekassedccoffeyrHCUPFactsFigures2007 |
| Proportion of toilets that are shared between \>1 patients | Constant | 60% | The remaining proportion (40%) of toilets were dedicated to one patient | N/A |
| Community colonisation rate | Uniform | $a = 0.01, b = 0.1$ |  | @crobachUnderstandingClostridiumDifficile2018 |
| Antimicrobial effect | Uniform | $a = 1.31, b = 1.87$ | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimingsAntibioticsHospitalacquiredClostridium2014 |
| CDI incidence | Uniform | $a=2.8, b=12.1$ | Per 10,000 patient-days | @marraIncidenceOutcomesAssociated2020 |
| Mean toilet frequenting rate | Uniform | $a=0.5,b=5$ | Per day; used as \lambda parameter for Poisson distribution |  |
| Mean toilet cleaning rate | Uniform | $a=0.5, b=3$ | Per day; used as \lambda parameter for Poisson distribution |  |
| Toilet contamination effect | Uniform | $a=0.01,b=0.99$ | Contamination impact by infected/colonised patients through toilet use |  |
| Toilet cleaning effect | Uniform | $a=0.01, b=0.99$ | Impact of toilet cleaning on contamination |  |

@@

A two-step approach to calibration was taken. Firstly, ABC was used to calibrate the model pre-outbreak data to identify posterior distributions for the parameters that recapitulated the baseline rate. Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009]. To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset as described in @sec-outbreak-generation. The priors used for ABC were selected to be compatible with increased transmission or colonisation rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency. To achieve this, latin hypercube sampling was repeated, and any parameters that did not satisfy these requirements (e.g., the parameter governing the antimicrobial prescription rate was _lower_ during the outbreak period) were discarded. At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures. The prior distributions for these parameters were selected to decrease transmission and colonisation -- generally the reverse of the 'outbreak' parameters. Again, only samples from the latin hypercube that satisfied these requirements were used.

@@

### Calibration of the \acr[first_use=TRUE]{SD} model

@@

To assess the sensitivity of the calibrated \acr{ABM} to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019]. Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) the mean delay between patient infection and transfer to a side room; 3) toilet cleaning rate; and 4) toilet cleaning effect. To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection \acr{ABC} Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value. The output metric was the number of cases of hospital-acquired infection/colonisation within the outbreak period. For each parameter value, the simulation was run with `r sens_n_seeds` different random number generator seeds to account for stochasticity in the model -- the output metric was the mean number of cases across the `r sens_n_seeds` runs.

@@

param_names <- c("toilet-frequenting-rate" = "Toilet frequenting rate",
                 "toilet-cleaning-rate" = "Toilet cleaning rate",
                 "toilet-contamination-effect" = "Toilet contamination effect",
                 "toilet-cleaning-effect" = "Toilet cleaning effect",
                 "community-colonisation-rate" = "Community colonisation rate",
                 "antibiotic-prescription-rate" = "Antibiotic prescription rate",
                 "antibiotic-effect" = "Antibiotic effect",
                 "random-colonisation" = "Random colonisation",
                 "side-room-delay-mean" = "Side room delay mean")

@@

```{r fig-posterior-densities, include=TRUE}
#| eval: true
#| echo: false
#| fig.width: 18
#| fig.height: 18
#| warning: false
#| fig-cap: "Changes in parameters during outbreak and enhanced infection control periods. Density plots show the posterior distributions of key simulation parameters after calibration using approximate Bayesian computation."
general_params <- abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(!contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type, linetype = type)) +
  geom_density(alpha = .3) +
  facet_wrap(~name, scales = "free_x") +
  theme_minimal() +
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "#66c2a5", "Outbreak" = "#fc8d62", "Control" = "#8da0cb")) +
  scale_linetype_manual(values = c("Baseline" = "solid", "Outbreak" = "dashed", "Control" = "dotted")) +
  # remove gridlines
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

@@

The sensitivity analysis explored the impact of different infection control practices on the number of cases observed during the outbreak period -- the results are shown in @fig-sensitivity-analysis. The analysis showed that the cleaning of toilets had a dramatic impact on the number of cases observed during the outbreak period, although the benefit appeared to plateau beyond a certain amount -- for example, cleaning toilets more than twice per day appeared to have marginal benefits. The number of hospital acquired cases did not appear to be sensitive to the antimicrobial prescription rate during the outbreak period. The cohorting of patients into single-occupancy rooms after colonisation/infection had an almost linear relationship with the number of cases.

@@

```{r}
# compare ABM community colonisation rate to SD model p
mean_sims %>% 
  summarise(mean = mean(`community-colonisation-rate`),
            sd = sd(`community-colonisation-rate`))
```

@@

@@

The model was designed to simulate the spread of infection/colonisation at a baseline (or background) rate, followed by an outbreak period of increased transmission, in turn followed by a period of enhanced infection control to return back to the baseline rate. The model was calibrated to a report of a _C. difficile_ outbreak reported by @salgadoAnalysisOutbreakClostridium2009. Approximate Bayesian computation was used to calibrate the model to the rates of infection reported in the paper. When model inputs were drawn from the posterior distributions, the model was able to simulate plausible outbreak scenarios that were similar to the observed rates. A sensitivity analysis was performed over parameters that represent infection control interventions (such as the frequency of toilet cleaning), demonstrating the possible applications of this model for informing operational infection control practices. Interventions that controlled the spread of infection, such as the cohorting of patients to single-occupancy beds, and the increased frequency of toilet cleaning, led to a reduced number of cases during the outbreak period.

To simulate outbreaks and periods of enhanced infection control, the model was designed to change parameters at discrete time points. For example, the rate of antimicrobial prescriptions was higher during the outbreak period, promoting the spread of infection and colonisation. Later, the prescription rate was reduced to help control the outbreak. The NetLogo model implementation was designed so that a fixed parameter set was used as an input, and the outbreak and enhanced IPC parameters were only used during their respective periods. In effect, all input parameters were provided as a fixed input -- some parameters were only used by the simulation at defined discrete intervals. This approach may be unsuitable for modelling of transmission without a clear outbreak pattern -- more complex implementations of time-varying parameters could be explored.

Although agent-based models have been shown to be useful for simulating and modelling infectious disease transmission, they have limitations. The model described here is a simplification of what is likely to be a highly intricate and complicated system. For example, transmission was assumed to be 'random' or acquired through the use of a contaminated toilet. In reality, other transmission pathways likely exist, such as through contact with staff or contaminated bedding. Antimicrobial exposure was considered to be a binary risk factor for an agent becoming infected or colonised. Antimicrobial class and duration of exposure are factors that are not accounted for in this model. An agent-based model that captures all the details that influence the spread of infection in a hospital setting is unlikely to be feasible, due to the computational challenge and the large number of parameters that would need to be estimated. A reasonable approach is for the model to be as simple as possible, while still capturing the features required to answer the research question [@sunSimpleComplicatedAgentbased2016]. Another limitation of agent-based models is the requirement for detailed data to calibrate and validate the model. Here, the data used was primarily from a retrospective descriptive dataset. As such, information on certain parameters was not available (e.g., the proportion of patients that were administered an antimicrobial during their admission) -- these parameters were estimated from the literature. An important future direction would be to prospectively inform and validate the model with data collected specifically for this purpose.

>>>>

---
title: Gastro-intestinal transmission of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
filters:
  - acronyms
acronyms:
  keys:
  - shortname: CDI
    longname: _Clostridium difficile_ infection
    markdown: longname
  - shortname: VRE
    longname: vancomycin-resistant _Enterococcus_
    markdown: longname
  - shortname: ESBL
    longname: extended-spectrum beta-lactamase
  - shortname: CPE
    longname: carbapenemase-producing Enterobacterales
  - shortname: WHO
    longname: World Health Organization
  - shortname: IPC
    longname: infection prevention and control
  - shortname: ABM
    longname: agent-based model
  - shortname: MRSA
    longname: methicillin-resistant _Staphylococcus aureus_
    markdown: longname
  - shortname: SARS-CoV-2
    longname: severe acute respiratory syndrome coronavirus 2
  - shortname: ABC
    longname: approximate Bayesian computation
  - shortname: IQR
    longname: interquartile range
  - shortname: FIFO
    longname: first-in, first-out
  - shortname: SD
    longname: system dynamics
  insert_loa: false
  insert_links: false
---

@@

\acr[case=sentence]{IPC} measures are designed to prevent the spread of these pathogens within healthcare settings. Commonly used interventions include hand hygiene, environmental cleaning, and the isolation of infected patients [@worldhealthorganizationGlobalReportInfection2022].
A key role of \acr{IPC} teams is to prepare for, declare, investigate, and control outbreaks of healthcare-associated infections.
In practice, an outbreak is declared after rates of infection have been compared to historical rates, and this is followed by a series of interventions to control the outbreak.
These interventions may include increased frequency and thoroughness of environmental cleaning, cohorting of infected patients (particularly into single-occupancy _en suite_ rooms), and increased hand hygiene compliance.
The impact of these interventions is often difficult to quantify.
Although evidence suggests that the interventions can help control outbreaks, studies almost always report their findings after a suite of interventions have been implemented -- therefore, the relative importance of each intervention is not usually quantifiable.
Furthermore, studies report findings from a population that may be very different from the outbreak population being managed -- differences can range from patient factors such as co-morbidities and rates of antimicrobial resistance, to institution practices such as rates of antimicrobial usage, and to structural differences such as hospital layout.

@@

The use of computer simulation techniques such as \acr{ABM} and can be used to model the spread of infectious diseases by constructing a virtual environment (such as a hospital), within which agents (such as patients and/or staff) interact. Agent-based modelling has been used to describe the transmission of \acr{MRSA} [@jaramilloAgentBasedModel2015] and \acr{SARS-CoV-2} [@kerrCovasimAgentbasedModel2021]. An advantage of \acr{ABM} over other simulation approaches is that models can be used to project the impact of interventions on transmission, prior to their implementation [@cassidyMathematicalModellingHealth2019]. This can be particularly useful for both planning and active management of outbreaks, where the impact of individual interventions is difficult to quantify [@cassidyMathematicalModellingHealth2019]. Another simulation approach is stochastic \acr{SD} modelling, which is a high-level approach that models the flow of patients between different states, such as susceptible and infected states, and is particularly useful for understanding feedback loops within systems.

Here, we describe an simulation model that models the transmission of faecal-oral-transmitted pathogens under antimicrobial selection pressure. Our aim was to explore the impact of various infection control strategies and interventions, particularly during outbreak periods. The parameters were specifically estimated using data from a \acr{CDI} outbreak reported by @salgadoAnalysisOutbreakClostridium2009, although our intention is that the approach is readily adaptable to other related antimicrobial-resistant pathogens such as \acr{VRE}, \acr{ESBL}, and \acr{CPE}. We compare an \acr{ABM} and a stochastic \acr{SD} model, and explore the advantages and disadvantages of each approach.

@@

-   $r_0$ is the baseline/*de novo* risk of infection,

-   $m$ is the increased (or decreased) relative risk of infection due to antimicrobial therapy,

@@

-   $dS/dt$ is the rate of change of susceptible/non-infected inpatients,
-   $S$ is the number of susceptible inpatients,
-   $I$ is the number of infected inpatients,
-   $A$ is the rate of new admissions,
-   $p$ is the community infection rate,
-   $\beta$ is the transmission coefficient,
-   $r$ is the rate of random (*de novo*) infection,
-   $D$ is the rate of patient discharges.

@@

```{r fig-sd-model, include=TRUE}
#| echo: false
#| fig.cap: "System dynamics model. $S$ and $I$ are the number of susceptible and infected patients, respectively; $A$ and $D$ are the rates of admitted and discharged patients, respectively; $p$ is the community infection rate; $\\beta$ is the transmission coefficient; and $r$ is the rate of random (*de novo*) infection."
knitr::include_graphics("./images/sd_model.svg")
```

@@

$$
D_{S,t} = \left \lfloor D_{\text{total}, t} \cdot \frac{S_t}{S_t + I_t} \right \rceil,
$$

@@

$$
D_{I,t} = \left \lfloor D_{\text{total}, t} \cdot \frac{I_t}{S_t + I_t} \right \rceil = D_{\text{total}, t} - D_{S,t},,
$$

@@

A similar approach was used to generate outbreaks in the stochastic system dynamics model, with the difference that the parameters were changed at: 1) onset of outbreak; and 2) end of outbreak. The parameters that were changed were: 1) $\beta$ (transmission coefficient); and 2) $p$ (community infection rate). The parameter $r$ (rate of random infection) was kept constant throughout the simulation.

@@

Model input parameters were broadly grouped into: 1) toilet (e.g., number of times toilet used per day, level of contamination after use or cleaning); 2) admission (e.g., community infection rate, antimicrobial prescription rate); and 3) infection parameters (e.g., relative risk increase of antimicrobial therapy on infection).

@@

| Name | Distribution | Parameters | Description | Reference |
|---------------|---------------|---------------|---------------|---------------|
| Baseline (pre-outbreak) antimicrobial prescription rate | Constant | 49.9% | Proportion of patients administered antimicrobials during admission | @versportenAntimicrobialConsumptionResistance2018 |
| Admission duration | Poisson | $\lambda = 4.6  \text{ days}$ |  | @levitkryankelixhauserastrangesekassedccoffeyrHCUPFactsFigures2007 |
| Proportion of toilets that are shared between \>1 patients | Constant | 60% | The remaining proportion (40%) of toilets were dedicated to one patient | N/A |
| Community infection rate | Uniform | $a = 0.01, b = 0.1$ |  | @crobachUnderstandingClostridiumDifficile2018 |
| Antimicrobial effect | Uniform | $a = 1.31, b = 1.87$ | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimingsAntibioticsHospitalacquiredClostridium2014 |
| CDI incidence | Uniform | $a=2.8, b=12.1$ | Per 10,000 patient-days | @marraIncidenceOutcomesAssociated2020 |
| Mean toilet frequenting rate | Uniform | $a=0.5,b=5$ | Per day; used as \lambda parameter for Poisson distribution |  |
| Mean toilet cleaning rate | Uniform | $a=0.5, b=3$ | Per day; used as \lambda parameter for Poisson distribution |  |
| Toilet contamination effect | Uniform | $a=0.01,b=0.99$ | Contamination impact by infected/colonised patients through toilet use |  |
| Toilet cleaning effect | Uniform | $a=0.01, b=0.99$ | Impact of toilet cleaning on contamination |  |

@@

A two-step approach to calibration was taken. Firstly, ABC was used to calibrate the model pre-outbreak data to identify posterior distributions for the parameters that recapitulated the baseline rate. Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgadoAnalysisOutbreakClostridium2009]. To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset as described in @sec-outbreak-generation. The priors used for ABC were selected to be compatible with increased transmission or infection rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency. To achieve this, latin hypercube sampling was repeated, and any parameters that did not satisfy these requirements (e.g., the parameter governing the antimicrobial prescription rate was _lower_ during the outbreak period) were discarded. At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures. The prior distributions for these parameters were selected to decrease transmission and infection -- generally the reverse of the 'outbreak' parameters. Again, only samples from the latin hypercube that satisfied these requirements were used.

@@

### Calibration of \acr[first_use=TRUE]{SD} model

@@

To assess the sensitivity of the calibrated \acr{ABM} to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019]. Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) the mean delay between patient infection and transfer to a side room; 3) toilet cleaning rate; and 4) toilet cleaning effect. To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection \acr{ABC} Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value. The output metric was the number of cases of hospital-acquired infection within the outbreak period. For each parameter value, the simulation was run with `r sens_n_seeds` different random number generator seeds to account for stochasticity in the model -- the output metric was the mean number of cases across the `r sens_n_seeds` runs.

@@

param_names <- c("toilet-frequenting-rate" = "Toilet frequenting rate",
                 "toilet-cleaning-rate" = "Toilet cleaning rate",
                 "toilet-contamination-effect" = "Toilet contamination effect",
                 "toilet-cleaning-effect" = "Toilet cleaning effect",
                 "community-colonisation-rate" = "Community infection rate",
                 "antibiotic-prescription-rate" = "Antibiotic prescription rate",
                 "antibiotic-effect" = "Antibiotic effect",
                 "random-colonisation" = "Random infection",
                 "side-room-delay-mean" = "Side room delay mean")

@@

```{r fig-posterior-densities, include=TRUE}
#| eval: true
#| echo: false
#| fig.width: 18
#| fig.height: 18
#| warning: false
#| fig-cap: "Changes in parameters during outbreak and enhanced infection control periods. Density plots show the posterior distributions of key simulation parameters after calibration using approximate Bayesian computation."
general_params <- abc_params_outbreak_control$unadj.values %>% 
  as_tibble %>% 
  dplyr::select(-`outbreak-start`, -`outbreak-end`, -`control-end`,
                -`antibiotic-effect`,
                -`random-colonisation`) %>% 
  dplyr::select(!contains("toilet")) %>% 
  pivot_longer(cols=everything()) %>%
  mutate(type = case_when(
    str_starts(name, "o-") ~ "Outbreak",
    str_starts(name, "c-") ~ "Control",
    .default = "Baseline")) %>%
  mutate(name = str_remove(name, "^o-|^c-")) %>%
  mutate(name = str_replace_all(name, "-", " ")) %>%
  mutate(name = str_replace_all(name, "colonisation", "infection")) %>%
  mutate(name = str_to_title(name)) %>% 
  # change type to ordered factor
  mutate(type = factor(type, levels = c("Baseline", "Outbreak", "Control"))) %>%
  ggplot(aes(x = value, ..scaled.., fill = type, linetype = type)) +
  geom_density(alpha = .3) +
  facet_wrap(~name, scales = "free_x") +
  theme_minimal() +
  # remove y axis
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  xlab("Parameter value") + 
  theme(legend.title = element_blank(),
        text=element_text(size=24)) + 
  # manually set colors
  scale_fill_manual(values = c("Baseline" = "#66c2a5", "Outbreak" = "#fc8d62", "Control" = "#8da0cb")) +
  scale_linetype_manual(values = c("Baseline" = "solid", "Outbreak" = "dashed", "Control" = "dotted")) +
  # remove gridlines
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

@@

The sensitivity analysis explored the impact of different infection control practices on the number of cases observed during the outbreak period -- the results are shown in @fig-sensitivity-analysis. The analysis showed that the cleaning of toilets had a dramatic impact on the number of cases observed during the outbreak period, although the benefit appeared to plateau beyond a certain amount -- for example, cleaning toilets more than twice per day appeared to have marginal benefits. The number of hospital acquired cases did not appear to be sensitive to the antimicrobial prescription rate during the outbreak period. The cohorting of patients into single-occupancy rooms after infection had an almost linear relationship with the number of cases.

@@

@@

_The discussion needs to explore the different modelling approaches used (SD vs ABM) and why we eventually settled on the ABM as the main model. This is sort of captured in the results, but maybe it needs to be here as well?_

@@

The model was designed to simulate the spread of infection at a baseline (or background) rate, followed by an outbreak period of increased transmission, in turn followed by a period of enhanced infection control to return back to the baseline rate. The model was calibrated to a report of a _C. difficile_ outbreak reported by @salgadoAnalysisOutbreakClostridium2009. Approximate Bayesian computation was used to calibrate the model to the rates of infection reported in the paper. When model inputs were drawn from the posterior distributions, the model was able to simulate plausible outbreak scenarios that were similar to the observed rates. A sensitivity analysis was performed over parameters that represent infection control interventions (such as the frequency of toilet cleaning), demonstrating the possible applications of this model for informing operational infection control practices. Interventions that controlled the spread of infection, such as the cohorting of patients to single-occupancy beds, and the increased frequency of toilet cleaning, led to a reduced number of cases during the outbreak period.

To simulate outbreaks and periods of enhanced infection control, the model was designed to change parameters at discrete time points. For example, the rate of antimicrobial prescriptions was higher during the outbreak period, promoting the spread of infection. Later, the prescription rate was reduced to help control the outbreak. The NetLogo model implementation was designed so that a fixed parameter set was used as an input, and the outbreak and enhanced IPC parameters were only used during their respective periods. In effect, all input parameters were provided as a fixed input -- some parameters were only used by the simulation at defined discrete intervals. This approach may be unsuitable for modelling of transmission without a clear outbreak pattern -- more complex implementations of time-varying parameters could be explored.

Although agent-based models have been shown to be useful for simulating and modelling infectious disease transmission, they have limitations. The model described here is a simplification of what is likely to be a highly intricate and complicated system. For example, transmission was assumed to be 'random' or acquired through the use of a contaminated toilet. In reality, other transmission pathways likely exist, such as through contact with staff or contaminated bedding. Antimicrobial exposure was considered to be a binary risk factor for an agent becoming infected. Antimicrobial class and duration of exposure are factors that are not accounted for in this model. An agent-based model that captures all the details that influence the spread of infection in a hospital setting is unlikely to be feasible, due to the computational challenge and the large number of parameters that would need to be estimated. A reasonable approach is for the model to be as simple as possible, while still capturing the features required to answer the research question [@sunSimpleComplicatedAgentbased2016]. Another limitation of agent-based models is the requirement for detailed data to calibrate and validate the model. Here, the data used was primarily from a retrospective descriptive dataset. As such, information on certain parameters was not available (e.g., the proportion of patients that were administered an antimicrobial during their admission) -- these parameters were estimated from the literature. An important future direction would be to prospectively inform and validate the model with data collected specifically for this purpose.
